### 2023-05-07 manuscript.tex
* manuscript.tex: Final edits.

Recently, studies of the so-called subregular classes, that is,
classes even weaker than regular relations, have proved a more
probable upper bound.  Such classes are conceptualized as a hierarchy,
not unlike the traditional Chomsky hierarchy in formal language
theory.  The remaining problem is how many phonological phenomena are
explainable and, more importantly, \textemph{not} explainable by each
class.  In this inquiry, tonal systems post an interesting challenge,
and thus require a more in-depth investigation.

Traditionally, tonal systems are accounted for by nonlinear phonology,
which involves the use of multiple phonological tiers and their
associations.  This process can be carefully separated into two ideas:
%
\begin{enumerate}
\item Tones can be independently represented on a separate tier just
  as segments do;
\item The associations between phonological tiers interact with tonal
  transformations in a nontrivial way.
\end{enumerate}
%
In this thesis, the two ideas are examined under the lens of attested
tonal systems found in Chinese languages with vastly different tonal
transformations.  Computationally, the former is embodied by linear
strictly local functions, and the latter by autosegmental strictly
local functions.  Their implication applies not only to tonal systems,
but also to the interfaces between linguistic levels in general.

@@

Specifically, the focus is placed on phonological
\textterm{transformations} as opposed to \textterm{phonotactics}.
Their distinction corresponds to the distinction between transducers
and acceptors in automata theory, which reflect translation and
membership problems respectively.  Loosely speaking, a transducer is a
function of type \(\alpha \to \beta\), while an acceptor is a function
of type \(\alpha \to \mathit{Boolean}\), that is, a
predicate.\footnote{Unless a gradient interpretation of phonotactics
  is required, in which case the output type can be, for example, an
  interval type.}  The latter will be briefly mentioned as the
theories are introduced.

@@

The discussion of \textterm{optimality theory} as in
\textcite{ps93otcigg} is intentionally omitted.  The reason is not
that optimality theory is a worse formalism, but merely that the
formalization of it results in computational devices out of the scope
of this thesis.  Specifically, optimality theory, due to its
constraint-optimizing nature, requires constraint optimizers rather
than transducers, which are potentially powerful enough to
overgenerate.  For such an implementation based on dynamic
programming, refer to \textcite{t95cot}.  For discussions of the
computational complexity of optimality theory, refer to
\textcite{e97egpot, i06sptotci, hkr09ecot}, among others.

@@

Attention should be paid to \enquote{linear}.  One interpretation is
that phonological transformations apply to a linear sequence of
segments.  This implies the assumption that segmental phonology, the
sort that is dealt with above, is disjoint from nonsegmental
phonology.  This assumption is computationally unsound, as
nonsegmental phonology does not necessarily require a
\enquote{nonlinear} account, and conversely segmental phonology may
require one.  Another interpretation has to do with the notion of
locality, which is an often misunderstood notion in theoretical
phonology.

@@

\subsection{Nonlinear Phonology}
Despite the remarks given above, nonsegmental phonology is precisely
what has inspired the development of nonlinear phonology, also known
as autosegmental phonology.  Early applications of nonlinear phonology
can be found in \textcite{c76vhngpam, g76ap}, among others.  Under
nonlinear phonology, phonological representations consist of multiple
associated tiers, each tier with its own linear sequence of terms.
For example, a syllable \textphon{ma} with a high level tone can be
represented as

@@

where \(\sigma\) stands for a syllable and H stands for a high tone.
Consequently, phonological transformations are generalized to operate
on associations.  An unbounded tone spreading rule is:

@@

An implicit assumption of it is that the rule applies iteratively from
left to right.  This intuition is, surprisingly, captured equally well
by linear transformations.  An equally acceptable formulation is the
linear transformation

@@

\ex. \(\exists A, \varphi, \psi\ldotp A \Rightarrow \varphi A \psi\)

where neither \(\varphi\) nor \(\psi\) is the identity term and
\(\Rightarrow\) consists of any number of rewrite steps.  Clearly,
phonological rules cannot express such a transformation.
\textcite{kk94rmprs} therefore suggests modeling phonological
transformations as \textterm{regular} relations.

Regular relations are relations whose domains and ranges are regular
languages.  An artificial regular relation, for example, is a relation
that rewrites any sequence that has an \textemph{even} number of
\(a\)s to a sequence with each \(a\) replaced by \(b\).  This is an
unattested phonological transformation, but it is a valid regular
relation, showcasing that regular relations are still computationally
too complex for modeling phonological transformations.

@@

Alternatively, regular relations can be represented as logical
transductions, specifically \textterm{monadic second-order} ones as
those of \textcite{eh01mdsttft}, which are actually able to represent
nonregular relations.\footnote{An example is the first-order definable
  total reduplication \(w \to ww\), whose output language is included
  in the class of tree-adjoining languages \parencite{sw94efecg}.}
Monadic second-order logic allows quantification over sets, which is
needed for the definition of general precedence \(<\) based on
immediate precedence \(\vartriangleleft\):\footnote{This is to assume
  the relations are nonrecursive.}

@@

This is understood as \enquote{\(x\) precedes \(y\) iff every
  transitive set with respect to immediate precedence that includes
  \(x\) also includes \(y\)}, which means the transitive closure also
includes \(y\).  A relation is then defined that expresses that an
\(a\) immediately precedes another \(a\) ignoring other non-\(a\)s
between them:

@@

The notion of the evenness of the number of \(a\)s is finally defined:

@@

Accordingly, the transduction is defined:\footnote{Observant readers
  may notice that an efficient transducer implementing said
  transduction reads the input \textemph{twice}, dispatching on the
  number of \(a\)s in the first pass.  This is due to the
  correspondence between monadic second-order logic and two-way
  finite-state transducers.}

@@

The concept of \textterm{strict locality} is defined for regular
languages, which concerns the recognition of substrings
\parencite{rp11apresh, rhfhlw13csc}.  Borrowing this concept, input
strictly local and output strictly local functions can be defined over
\textterm{linear} representations, which \enquote{remember} a finite
length of input and output substrings respectively
\parencite{ceh14lslsf}.  Similar to subsequential functions, output
strictly local functions can be further classified into left and right
variants \parencite{ceh15oslf}.

@@

Logical encodings enable the generalization of strict locality over
\textterm{autosegmental} representations \parencite{cj19aislf}.  Under
a predicate logic, associations are understood as a binary relation
\(\mathit{Assoc}\) where \(\mathit{Assoc}(a, b)\) indicates that \(a\)
and \(b\) are associated, given that \(a\) and \(b\) are on separate
tiers.\footnote{A question can be asked whether \(\mathit{Assoc}\) is
  symmetric.  Although this question has no particular bearing,
  \(\mathit{Assoc}\) is assumed to be antisymmetric to allow the
  interpretation of autosegmental representations as directed graphs.}
For example, the unbounded tone spreading rule can be defined as the
recursive logical transduction

@@

An alternative view of boundaries is \textterm{groupings}, which can
render the transformation nonregular.  Even if we ignore this nature
of arbitrary groupings by restricting self-embedding in some manner,
there is still the issue of determining whether two terms are in the
same group.  Suppose groupings are delimited by conventional brackets,
to determine whether two terms are in the same group, the transducer
potentially needs to traverse the whole input sequence.  This is more
apparent in the case of autosegmental representations, where groupings
are represented as represented as trees, which are a subset of
directed graphs.\footnote{Such trees with a limited depth are known as
  \enquote{prosodic hierarchy}, for which see \textcite{nv86pp}.} The
relation that decides whether two terms are in the same group is then

@@

show that certain tone 3 \textemph{optionally} changes to tone 2.  To
account for such contrasts, various analyses have been developed.  Two
such analyses, where one is representative for \enquote{syntactic}
analyses, and the other for \enquote{prosodic} analyses, will be
discussed.

\subsection{Syntactic Analysis}
\textcite{c00tspcd} argues for a analysis under which feet are built
based on syntactic structures, which in turn govern the application of
tone 3 sandhi.  As \textcite{d07psc} states, Although this analysis
includes a notion of feet, they are nonetheless not related to prosody
given their insensitivity to stress.  More crucially, the tone 3
sandhi rule makes reference to the syntactic structure when it is
optional.  Therefore, it is considered a \enquote{syntactic} analysis
here.

@@

Alternatively, we may consider that the trees be transformed to linear
sequences suitable as inputs to the later level.  This can be a simple
recursive process that flattens the input trees using an associative
operation that prefixes boundaries.\footnote{This process can also be
  represented as a strictly local tree transducer, given that linear
  sequences are isomorphic to single-pathed graphs, that is,
  unary-branching trees.  See \textcite{ioj20qtt, jh20isltt} for
  detailed discussions of such transducers.}  This tree transformation
can be nondeterministic to account for the optionality, but the level
it feeds, tone 3 sandhi, remains deterministic.\footnote{Under
  nondeterminism, an output sequence that contains a
  \(\check{\sigma}\check{\sigma}\) substring simply fails and
  backtracks.  Although \textcite{c00tspcd} reports several data that
  contain such substrings, they are more likely due to emphatic
  stresses.}  Considering only the nontrivial cases, we have

@@

We see how convoluted the analysis becomes once the phonological
process is no longer subregular.  Even though tone 3 sandhi is a one
of the simplest phonological processes, the \textemph{condition} it
applies heavily depends on syntactic information.  Still worse, its
application can be nondeterministic.  By distributing the workload to
tone 3 sandhi, its computational nature is effectively obscured.
Moreover, using autosegmental representations to represent
\textemph{hierarchical} structures is in some sense an abuse.

@@

\ex.
\a. \(\grave{\sigma}\grave{\sigma}\grave{\sigma} \to
\grave{\sigma}\check{\sigma}\grave{\sigma}\)
\hfill(right-to-left)
\b. \(\check{\sigma}\check{\sigma}\check{\sigma} \to
\check{\sigma}\acute{\sigma}\check{\sigma}\)
\hfill(right-to-left)
\b. \(\acute{\sigma}\acute{\sigma}\acute{\sigma} \to
\bar{\sigma}\bar{\sigma}\acute{\sigma}\)
\hfill(left-to-right)

There has been debates on how these rules interact and, by extension,
how to explain such interactions.  Rule-based analyses such as
\posscite{t87tstd, z87ptal} proposals rely on stipulations to account
for the directionalities as well as the feeding relations, therefore
offering no real explanation.  In the subregular zoo, instead of
directionalities, the difference in patterns amounts to the difference
in logical formulae, that is, the computational nature of the rules.
The problem is then how to also accommodate the feeding relations.  A
relevant \enquote{boundaryless} computational analysis will be
discussed before a new \enquote{boundaryful} analysis is presented.

\subsection{Bounadaryless Analysis}
\posscite{c18catsi} analysis is based on the observation that the
three strictly local functions can be combined into a single one,
which must remember a pair of input and output substrings.  It is thus
an input-output strictly local function.  Following our convention so
far, the idea will be described using logical formulae instead of
transducers.  The logical transduction, without consideration of the
feeding relations, is as follows:

@@

that appears as if every bracket is rewritten into a boundary term.
In other words, a left-branching structure causes the substring of
first two syllables to undergo tone sandhi, while a right-branching
structure causes the substring of last two syllables to undergo tone
sanhi, when boundaries are outputted at all.

As usual, the tree transformation is nondeterministic, so that
boundaries are allowed to be omitted.  With this in mind, it is still
true that the patterns that are attested for \textemph{both}
left-branching and right-branching structures can be used to justify
the trisyllabic generalization, as these patterns occur when
boundaries are omitted.  The transduction therefore remains strictly
local, while the originally conflicting data can now be well
explained:

@@

To decide the feeding relations, data involving rule interactions are
examined.  LL sandhi bleeds FL sandhi:

\ex.
\a. \(\grave{\sigma}(\check{\sigma}\check{\sigma}) \to
\{\grave{\sigma}\#\acute{\sigma}\check{\sigma},
\grave{\sigma}\acute{\sigma}\check{\sigma}\}\)
\b. \((\grave{\sigma}\check{\sigma})\check{\sigma} \to
\{\bar{\sigma}\check{\sigma}\#\check{\sigma},
\grave{\sigma}\acute{\sigma}\check{\sigma}\}\)

FL sandhi counterbleeds FF sandhi:

\ex.
\a. \(\grave{\sigma}(\grave{\sigma}\check{\sigma}) \to
\{\grave{\sigma}\#\bar{\sigma}\check{\sigma},
\check{\sigma}\bar{\sigma}\check{\sigma}\}\)
\b. \((\grave{\sigma}\grave{\sigma})\check{\sigma} \to
\{\check{\sigma}\grave{\sigma}\#\check{\sigma},
\check{\sigma}\bar{\sigma}\check{\sigma}\}\)

LL sandhi feeds RR sandhi:

\ex.
\a. \(\acute{\sigma}(\check{\sigma}\check{\sigma}) \to
\{\acute{\sigma}\#\acute{\sigma}\check{\sigma},
\bar{\sigma}\acute{\sigma}\check{\sigma}\}\)
\b. \((\acute{\sigma}\check{\sigma})\check{\sigma} \to
\{\acute{\sigma}\check{\sigma}\#\check{\sigma},
\bar{\sigma}\acute{\sigma}\check{\sigma}\}\)

FF sandhi feeds LL sandhi:

\ex.
\a. \(\check{\sigma}(\grave{\sigma}\grave{\sigma}) \to
\{\check{\sigma}\#\check{\sigma}\grave{\sigma},
\acute{\sigma}\check{\sigma}\grave{\sigma}\}\)
\b. \((\check{\sigma}\grave{\sigma})\grave{\sigma} \to
\{\check{\sigma}\grave{\sigma}\#\grave{\sigma},
\acute{\sigma}\check{\sigma}\grave{\sigma}\}\)

@@

\section{Interim Summary}
This section has argued for a modular analysis of tonal transformation
that is sensitive to the hierarchical structure.  The point is not to
suggest that a certain analysis is inherently better than another.
Rather, it argues that such a modular analysis is computationally
feasible and clarifies the boundaries between linguistic levels.  Even
if the overall complexity of the composed system remains the same, a
modular analysis allows for the local reasoning of each module due to
abstraction and encapsulation.  This is also why modularization is
heavily studied in programming language theory.

@@

The transformation then simply rewrites all \(\acute{a}\)s to \(a\)s
and all \(\grave{a}\)s to \(b\)s.  All is well, except now the
representation not only encodes the evenness information, but also
renders many sequences invalid.  It is only reasonable when it serves
as an \textemph{intermediate} representation that encodes the decision
of the evenness.  Even then, the decision would have had the same
computational complexity as before anyway.

@@

It is worth it to ask whether autosegmental representations are really
needed.  For example, it is not apparent why the excessive terms on
the tonal tier should be present at all if they always end up unused
by the transformation.  The problem is that the tonal status of a
specific syllable is always tied to the association relation, forcing
us to retreat to this unnecessary indirection and obscuring the nature
of left dominance.

@@

This undesirably imposes a further assumption of the
\enquote{underlying} association, even though it never surfaces in
this specific mapping.  Supposedly, the assumption is made to account
for the trivial case of monosyllabic sequences, where the mapping is

@@

This transduction directly corresponds to the theoretical analysis and
assumes nothing about the underlying association.  This gives rise to
the question whether the \enquote{garbage} tonal terms are retained in
the surface forms, if a normalization procedure is in effect.
Intuitively, the surface forms should always be normalized to encode
the \textemph{least} information as possible.  Such normalization can
be achieved by tailoring the domain predicate that decides whether or
not a term is copied:\footnote{And, accordingly, the predecessor and
  successor functions are adjusted to the domain.}

@@

Another more serious problem is that this transformation misses the
\textform{yángrù} pattern.  This pattern occurs when the first
syllable both starts with a voiced onset (\enquote{\textform{yáng}})
and ends with a stop coda (\enquote{\textform{rù}}), and is much
different from the above pattern:

@@

Intuitively, it spreads the first tonal term to all syllables and
associates the second tonal term to the last syllable.

@@

\subsection{Linear Analysis}
At this point, it is worth it to revisit the linear transformation.
As we notice, there are only \textemph{two} possible configurations in
the non-\textform{yángrù} pattern.  This results in a correspondence
to the previously assumed linear representations:

@@

The above transformation, like its autosegmental counterpart, is
nonrecursive.  However, the \textform{yángrù} pattern still requires a
recursive definition, given that it is characteristically an unbounded
tone spreading rule:

@@

The combination of the two transformations, again, requires the
distinction between the \textform{yángrù} pattern and
non-\textform{yángrù} patterns.  The same technique can be adapted to
the linear transformation, with the caveat that tone spreading applies
up to the \textemph{penultimate} syllable.  The resulting transduction
is as follows, omitting predicates that carry their original
definitions:\footnote{For convenience, \textform{yángrù} is assumed to
  be a unique tone in the inputs, although such an assumption does not
  otherwise change the overall complexity.}

@@

where \(\mathit{Rel}_{\mathit{Mor}}\) causes the unfortunate presence
of existential quantification.  In \textcite{z20eaislfmdft}, this
existential quantification is left out from the transduction,
resulting in the mischaracterization of it as strictly local.  This
cannot be true as the transduction must be logically closed, while the
removal of the existential quantification leads to a free variable.
Regardless, this transduction produces the following mappings:

@@

As a last resort, we can consider how to preserve the intuition of the
transduction while removing the unnecessary assumptions.  An immediate
notice to make is that the addition of \(\mathit{Rel}_{\mathit{Mor}}\)
contributes nothing to the transduction whatsoever.  It is introduced
purely on a theoretical basis, much like what we have seen in Shanghai
Wu.  Crucially, it disambiguates the cases where a floating tone can
be either from the first or second syllable:

@@

Nonetheless, there is the possibility that such a distinction is
simply \textemph{unfeasible}, given that existential quantification is
needed to make use of the morphemic information.  In fact, it is
rather reminiscent of the problem of grouping discussed above.  At the
very least, this scenario does not appear in Suzhou Wu, not even when
\enquote{light}-initial sequences are concerned, although they will
not be the focus here.

Another problem is that the transduction is forced to assume that the
second tonal term for exceptional patterns must always be underlyingly
associated with the first syllable.  This assumption is acceptable as
a phonotactic condition, but it does raise the question whether it can
be dropped.  Overall, the modified transduction is defined as

>>>>

Recently, studies of the so-called subregular classes, that is,
classes even weaker than regular relations, have proved to be a more
probable upper bound.  Such classes are conceptualized as a hierarchy,
not unlike the traditional Chomsky hierarchy in formal language
theory.  The remaining problem is how many phonological phenomena are
explainable and, more importantly, \textemph{not} explainable by each
class.  In this inquiry, tonal systems post an interesting challenge,
and thus require a more in-depth investigation.

Traditionally, tonal systems are accounted for by nonlinear phonology,
which involves the use of multiple phonological tiers and their
associations.  This process can be carefully separated into two ideas:
%
\begin{enumerate}
\item Tones can be independently represented on a separate tier just
  as segments do;
\item The associations between phonological tiers can be manipulated
  in a nontrivial way.
\end{enumerate}
%
In this thesis, the two ideas are examined under the lens of attested
tonal systems found in Chinese languages with vastly different tonal
transformations.  Their implication applies not only to tonal systems,
but also to the interfaces between linguistic levels in general.

@@

Specifically, the focus is placed on phonological
\textterm{transformations} as opposed to \textterm{phonotactics}.
Their distinction corresponds to the distinction between transducers
and acceptors in automata theory, which reflect translation and
membership problems respectively.  Loosely speaking, a transducer is a
function of type \(\alpha \to \beta \to \mathit{Boolean}\), while an
acceptor of type \(\alpha \to \mathit{Boolean}\), that is, a
predicate.\footnote{For deterministic transducers, the types are
  isomorphic to \(\alpha \to \beta\).}\footnote{Unless a gradient
  interpretation of phonology is required, in which case the output
  type can be, for example, an interval type.}  The latter will be
briefly mentioned as the theories are introduced.

@@

The discussion of \textterm{optimality theory} as in
\textcite{ps93otcigg} is intentionally omitted.  The reason is not
that optimality theory is a worse formalism, but merely that its
formalization results in computational devices out of the scope of
this thesis.  Specifically, optimality theory, due to its
constraint-optimizing nature, requires constraint optimizers rather
than transducers, which are potentially powerful enough to
overgenerate.  For such an implementation based on dynamic
programming, refer to \textcite{t95cot}.  For discussions of the
computational complexity of optimality theory, refer to
\textcite{e97egpot, i06sptotci, hkr09ecot}, among others.

@@

Attention should be paid to \enquote{linear}.  One interpretation is
that phonological transformations apply to a linear sequence of
segments.  This assumes that segmental phonology, the sort that is
dealt with above, is disjoint from nonsegmental phonology.  This
assumption is computationally unsound, as nonsegmental phonology does
not necessarily require a \enquote{nonlinear} account, and conversely
segmental phonology may require one.  Another interpretation has to do
with the notion of locality, which is an often misunderstood notion in
theoretical phonology.

@@

\subsection{Nonlinear Phonology}
Despite the remarks made above, nonsegmental phonology is precisely
what has inspired the development of nonlinear phonology, also known
as autosegmental phonology.  Early applications of nonlinear phonology
can be found in \textcite{c76vhngpam, g76ap}, among others.  Under
nonlinear phonology, phonological representations consist of multiple
associated tiers, each tier with its own linear sequence of terms.
For example, a syllable \textphon{ma} with a high level tone can be
represented as

@@

where \(\sigma\) stands for a syllable and H stands for a high tone.
Consequently, phonological transformations are generalized to operate
on associations.  An unbounded tone spreading rule can be given as:

@@

An implicit assumption is that the rule applies iteratively from left
to right.  This intuition is, surprisingly, captured equally well by
linear transformations.  An equally acceptable formulation is the
linear transformation

@@

\ex. \(\exists A, \varphi, \psi\ldotp A \to^{+} \varphi A \psi\)

where neither \(\varphi\) nor \(\psi\) is the identity term and
\(\to^{+}\) is the transitive closure of the rewrite relation.
Clearly, phonology does not permit such a structure.
\textcite{kk94rmprs} therefore suggests modeling phonological
transformations as \textterm{regular} relations.

Regular relations are relations whose domains and ranges are regular
languages.  An artificial regular relation, for example, is a relation
that rewrites any sequence that has an \textemph{even} number of \(a\)
to a sequence with each \(a\) replaced by \(b\).  This is an
unattested phonological transformation, but it is a valid regular
relation, showcasing that regular relations are still computationally
too complex for modeling phonological transformations.

@@

Alternatively, regular relations can be represented as logical
transductions, specifically \textterm{monadic second-order} ones as
those in \textcite{eh01mdsttft}, which are actually able to represent
nonregular relations.\footnote{An example is the first-order definable
  total reduplication \(w \to ww\), whose output language is included
  in the class of tree-adjoining languages \parencite{sw94efecg}.}
Monadic second-order logic allows quantification over sets, which is
needed for the definition of general precedence \(<\) based on
immediate precedence \(\vartriangleleft\):\footnote{This is to assume
  the relations are nonrecursive.}

@@

This is understood as \enquote{\(x\) precedes \(y\) iff every
  transitive set with respect to immediate precedence that includes
  \(x\) also includes \(y\)}, which means the transitive closure also
includes \(y\).  A relation is then defined that expresses that an
\(a\) immediately precedes another \(a\) ignoring any non-\(a\)
between them:

@@

The notion of the evenness of the number of \(a\) is finally defined:

@@

Accordingly, the transduction is defined:\footnote{Observant readers
  may notice that an efficient transducer implementing said
  transduction reads the input \textemph{twice}, dispatching on the
  number of \(a\) in the first pass.  This is due to the
  correspondence between monadic second-order logic and two-way
  finite-state transducers.}

@@

The concept of \textterm{strict locality} is defined for regular
languages, which concerns the recognition of substrings
\parencite{rp11apresh, rhfhlw13csc}.  Borrowing this concept, input
and output strictly local functions can be defined over
\textterm{linear} representations, which \enquote{remember} a finite
length of input and output substrings respectively
\parencite{ceh14lslsf}.  Similar to subsequential functions, output
strictly local functions can be further classified into left and right
variants \parencite{ceh15oslf}.

@@

Logical encodings enable the generalization of strict locality over
\textterm{autosegmental} representations \parencite{cj19aislf}.  Under
a predicate logic, associations are understood as a binary relation
\(\mathit{Assoc}\) where \(\mathit{Assoc}(a, b)\) indicates that \(a\)
and \(b\) are associated, where \(a\) and \(b\) are on separate
tiers.\footnote{A question can be asked whether \(\mathit{Assoc}\) is
  symmetric.  Although this question has no particular bearing,
  \(\mathit{Assoc}\) is assumed to be antisymmetric to allow the
  interpretation of autosegmental representations as directed graphs.}
For example, the unbounded tone spreading rule can be defined as the
recursive logical transduction

@@

An alternative view of boundaries is \textterm{groupings}, which can
render the transformation nonregular.  Even if we ignore this nature
of arbitrary groupings by restricting self-embedding in some manner,
there is still the issue of determining whether two terms are in the
same group.  Suppose groupings are delimited by conventional brackets,
to determine whether two terms are in the same group, the transducer
potentially needs to traverse the whole input sequence.  This is more
apparent in the case of autosegmental representations, where groupings
are represented as trees, which are a subset of directed
graphs.\footnote{Such trees with a limited depth are known as
  \enquote{prosodic hierarchy}, for which see \textcite{nv86pp}.} The
relation that decides whether two terms are in the same group is then

@@

show that certain tone 3 \textemph{optionally} changes to tone 2.  To
account for such contrasts, various analyses have been developed.  Two
such analyses will be discussed, where one is representative for
\enquote{syntactic} analyses and the other for \enquote{prosodic}
analyses.

\subsection{Syntactic Analysis}
\textcite{c00tspcd} argues for a analysis under which feet are built
based on syntactic structures, which in turn govern the application of
tone 3 sandhi.  As \textcite{d07psc} states, Although this analysis
includes a notion of feet, they are nonetheless not related to prosody
given their insensitivity to stress.  More crucially, the tone 3
sandhi rule makes reference to the syntactic structure for the
applicability.  Therefore, it is considered a \enquote{syntactic}
analysis here.

@@

Alternatively, we may consider that the trees be transformed to linear
sequences suitable as inputs to the later level.  This can be a simple
recursive process that flattens the input trees using an associative
operation that prefixes boundaries.\footnote{This process can also be
  represented as a strictly local tree transducer, as linear sequences
  are isomorphic to single-pathed graphs, that is, unary-branching
  trees.  See \textcite{ioj20qtt, jh20isltt} for detailed discussions
  of such transducers.}  This tree transformation can be
nondeterministic to account for the optionality, but the level it
feeds, tone 3 sandhi, remains deterministic.\footnote{Under
  nondeterminism, an output sequence that contains a
  \(\check{\sigma}\check{\sigma}\) substring simply fails and
  backtracks.  Although \textcite{c00tspcd} reports several data that
  contain such substrings, they are more likely due to emphatic
  stresses.}  Considering only the nontrivial cases, we have

@@

We see how convoluted the analysis becomes once the phonological
process is no longer subregular.  Even though tone 3 sandhi is one of
the simplest phonological processes, the \textemph{condition} it
applies heavily depends on syntactic information.  Still worse, its
application can be nondeterministic.  By distributing the workload to
tone 3 sandhi, its computational nature is effectively obscured.
Moreover, using autosegmental representations to represent
\textemph{hierarchical} structures is in some sense an abuse.

@@

\ex.
\a. \(\grave{\sigma}\grave{\sigma}\grave{\sigma} \to
\grave{\sigma}\check{\sigma}\grave{\sigma}\)
\hfill(right-to-left)
\b. \(\check{\sigma}\check{\sigma}\check{\sigma} \to
\check{\sigma}\acute{\sigma}\check{\sigma}\)
\hfill(right-to-left)
\b. \(\acute{\sigma}\acute{\sigma}\acute{\sigma} \to
\bar{\sigma}\acute{\sigma}\acute{\sigma} \to
\bar{\sigma}\bar{\sigma}\acute{\sigma}\)
\hfill(left-to-right)

There has been debates on how these rules interact and, more
importantly, how to explain such interactions.  Rule-based analyses
such as \posscite{t87tstd, z87ptal} proposals rely on stipulations to
account for the directionalities as well as the feeding relations,
therefore offering no real explanation.  In the subregular zoo,
instead of directionalities, the difference in patterns amounts to the
difference in logical formulae, that is, the computational nature of
the rules.  The problem is then how to also accommodate the feeding
relations.  A relevant \enquote{boundaryless} computational analysis
will be discussed before a new \enquote{boundaryful} analysis is
presented.

\subsection{Bounadaryless Analysis}
\posscite{c18catsi} analysis is based on the observation that the
three strictly local functions can be combined into a single one,
which must remember a pair of input and output substrings.  It is thus
an input--output strictly local function.  Following our convention so
far, the idea will be described using logical formulae instead of
transducers.  The logical transduction, without consideration of the
feeding relations, is as follows:

@@

that appears as if every bracket is rewritten into a boundary term.
In other words, when boundaries are outputted at all, a left-branching
structure causes the substring of first two syllables to undergo tone
sandhi, while a right-branching one does that of last two.  As usual,
the tree transformation is nondeterministic so that boundaries are
allowed to be omitted.

With this in mind, it is still true that the patterns that are
attested for \textemph{both} left-branching and right-branching
structures can be used to justify the trisyllabic generalization, as
these patterns occur when boundaries are omitted.  The transduction
therefore remains strictly local, while the originally conflicting
data can now be well explained:

@@

To decide the feeding relations, data involving rule interactions are
examined:

\ex.
\a. LL sandhi bleeds FL sandhi
\a. \(\grave{\sigma}(\check{\sigma}\check{\sigma}) \to
\{\grave{\sigma}\#\acute{\sigma}\check{\sigma},
\grave{\sigma}\acute{\sigma}\check{\sigma}\}\)
\b. \((\grave{\sigma}\check{\sigma})\check{\sigma} \to
\{\bar{\sigma}\check{\sigma}\#\check{\sigma},
\grave{\sigma}\acute{\sigma}\check{\sigma}\}\)
%
\z.
\b. FL sandhi counterbleeds FF sandhi
\a. \(\grave{\sigma}(\grave{\sigma}\check{\sigma}) \to
\{\grave{\sigma}\#\bar{\sigma}\check{\sigma},
\check{\sigma}\bar{\sigma}\check{\sigma}\}\)
\b. \((\grave{\sigma}\grave{\sigma})\check{\sigma} \to
\{\check{\sigma}\grave{\sigma}\#\check{\sigma},
\check{\sigma}\bar{\sigma}\check{\sigma}\}\)
%
\z.
\b. LL sandhi feeds RR sandhi
\a. \(\acute{\sigma}(\check{\sigma}\check{\sigma}) \to
\{\acute{\sigma}\#\acute{\sigma}\check{\sigma},
\bar{\sigma}\acute{\sigma}\check{\sigma}\}\)
\b. \((\acute{\sigma}\check{\sigma})\check{\sigma} \to
\{\acute{\sigma}\check{\sigma}\#\check{\sigma},
\bar{\sigma}\acute{\sigma}\check{\sigma}\}\)
%
\z.
\b. FF sandhi feeds LL sandhi
\a. \(\check{\sigma}(\grave{\sigma}\grave{\sigma}) \to
\{\check{\sigma}\#\check{\sigma}\grave{\sigma},
\acute{\sigma}\check{\sigma}\grave{\sigma}\}\)
\b. \((\check{\sigma}\grave{\sigma})\grave{\sigma} \to
\{\check{\sigma}\grave{\sigma}\#\grave{\sigma},
\acute{\sigma}\check{\sigma}\grave{\sigma}\}\)

@@

\section{Interim Summary}
This section has argued for a modular analysis of tonal transformation
that is sensitive to the hierarchical structure.  The point is not to
suggest that a certain analysis is inherently better than another.
Rather, it argues that such a modular analysis is computationally
feasible and clarifies the boundaries between linguistic levels.  Even
if the overall complexity of the composed system remains the same, a
modular analysis allows for the local reasoning of each module due to
abstraction and encapsulation.

@@

The transformation then simply rewrites each \(\acute{a}\) to \(a\)
and \(\grave{a}\) to \(b\).  All is well, except now the
representation not only encodes the evenness information, but also
renders many sequences invalid.  It is only reasonable when it serves
as an \textemph{intermediate} representation that encodes the decision
of the evenness.  Even then, the decision would have had the same
computational complexity as before anyway.

@@

It is worth it to ask whether autosegmental representations are really
needed.  For example, it is not apparent why the excessive terms on
the tonal tier should be present at all if they always end up unused
by the transformation.  A bigger problem is that the tonal status of a
specific syllable is always tied to the association relation, forcing
us to retreat to this unnecessary indirection and obscuring the nature
of left dominance.

@@

This imposes an undesirable assumption of the \enquote{underlying}
association, even though it never surfaces in this specific mapping.
Supposedly, the assumption is made to account for the trivial case of
monosyllabic sequences, where the mapping is

@@

This transduction directly corresponds to the theoretical analysis and
assumes nothing about the underlying association.  This gives rise to
the question whether the \enquote{garbage} tonal terms are retained in
the surface forms---if a normalization procedure is in effect.
Intuitively, the surface forms should always be normalized to encode
the \textemph{least} information as possible.  Such normalization can
be achieved by tailoring the domain predicate that decides whether or
not a term is copied:\footnote{And, accordingly, the predecessor and
  successor functions are adjusted to the domain.}

@@

A more severe problem is that this transformation misses the
\textform{yángrù} pattern.  This pattern occurs when the first
syllable both starts with a voiced onset (\enquote{\textform{yáng}})
and ends with a stop coda (\enquote{\textform{rù}}), and is much
different from the above pattern:

@@

Effectively, it spreads the first tonal term to all syllables and
associates the second tonal term to the last syllable.

@@

\subsection{Linear Analysis}
At this point, it is worth it to revisit the linear transformation.
As we notice, there are only \textemph{two} possible configurations in
the non-\textform{yángrù} patterns.  This results in a correspondence
to the previously assumed linear representations:

@@

The above transformation, like its autosegmental counterpart, is
nonrecursive.  However, the \textform{yángrù} pattern still requires a
recursive definition given that it is characteristically a tone
spreading rule:

@@

The combination of the two transformations, again, requires the
distinction between the \textform{yángrù} and non-\textform{yángrù}
patterns.  The same technique can be adapted to the linear
transformation, with the caveat that tone spreading applies up to the
\textemph{penultimate} syllable.  The resulting transduction is as
follows, omitting predicates that carry their original
definitions:\footnote{For convenience, \textform{yángrù} is assumed to
  be a unique tone in the inputs, although such an assumption does not
  otherwise change the overall complexity, as \textform{yángrù} can be
  recognized by the segmental features.}

@@

where \(\mathit{Rel}_{\mathit{Mor}}\) causes the unfortunate presence
of existential quantification.  In \textcite{z20eaislfmdft}, this
existential quantification is left out from the transduction,
resulting in its mischaracterization as strictly local.  This cannot
be true, as the transduction must be logically closed, while the
removal of the existential quantification leads to a free variable.
Regardless, this transduction produces the following mappings:

@@

As the very least, we can consider how to preserve the intuition of
the transduction while removing the unnecessary assumptions.  An
immediate note to make is that the addition of
\(\mathit{Rel}_{\mathit{Mor}}\) contributes nothing to the
transduction whatsoever.  It is introduced purely on a theoretical
basis, much like what we have seen in the case of Shanghai Wu.
Crucially, it disambiguates the cases where a floating tone can be
either from the first or second syllable:

@@

Nonetheless, there is the possibility that such a distinction is
simply \textemph{unfeasible} given that existential quantification is
needed to make use of the morphemic information.  In fact, it is
rather reminiscent of the problem of grouping previously discussed.
Moreover, this scenario does not appear in Suzhou Wu, not even when
\enquote{light}-initial sequences are concerned.

Another problem is that the transduction is forced to assume that the
relevant tonal terms for exceptional patterns must always be
underlyingly associated with the first syllable.  This assumption is
acceptable as a phonotactic condition, but it does raise the question
whether it can be dropped.  Overall, the modified transduction is
defined as
